Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2016-12-15 paper/numerical_experiments_v4_cmame.tex
added missing abstract
minor edits

\Cref{tab:ref3D_diffmesh} shows the error at the end of each adaptive iteration. Each iteration of the adaptive algorithm uses the solution of the previous iteration as its initial guess. The number of degrees of freedom of each of the primary (and, if applicable, auxiliary) variables at each iteration is also given; the supplementary adjoint is solved on the high-fidelity mesh and thus each of its variables has the same number of degrees of freedom as each of the primary variables in the high-fidelity inverse problem. The high-fidelity inverse problem does not converge when the low-fidelity solution is used as an initial guess. Instead, we solve the inverse problem for an intermediate model
%
\begin{equation}
\nabla\cdot(n\vec{V}u - nD\nabla u) + k_ru^2 = f(q)
\end{equation}
%
and use the solution as an initial guess for the high-fidelity problem; equivalently, we use two steps of natural continuation on the reaction parameter: $k_r=0$, then $k_r=4.2\cdot10^{-4}$.
%
\begin{table}[htbp]
\caption{Runtime and relative errors of adaptive algorithm iterations given relative error tolerance of $0.1\%$; relative errors are given with respect to the high-fidelity QoI estimate.}
\label{tab:ref3D_diffmesh}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{DOFs} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative \\
& & & & (Estimated) & (Actual) & Error (Actual)  \\ \hline
LF   & 0    & 9537  & 168710 & -16463 & -85663 & -103    \\
MF01 & 5.2  & 13417 & 167366 & -7207  & -84319 & -102    \\
MF02 & 11.4 & 17895 & 89777  & -6208  & -6730  & -8.10   \\
MF03 & 16.3 & 21001 & 85880  & -2473  & -2833  & -3.41   \\
MF04 & 22.0 & 24528 & 83902  & -711   & -855   & -1.03   \\
MF05 & 27.7 & 27984 & 83119  & 32     & -72    & -0.087  \\
HF   & 100  & 70785 & 83047  & --     & --     & --    \\ \hline
\end{tabular}
\end{table}
%

@@

To illustrate this offline-online application, we generate ten sets of noisy observations to represent the actual data gained during the online phase; the noisy observations are generated by taking the observations used in the adaptive algorithm and adding Gaussian white noise with standard deviation of $\sigma=0.02$ (equivalent to, on average, $5\%$ of the observed values). We then solve the inverse problem using each of the mixed-fidelity models depicted in \Cref{tab:ref3D_diffmesh,fig:divvy3D_diffmesh} as well as the high-fidelity model. The low-fidelity inverse problem is first solved and used as an initial guess for the higher-fidelity problems; we note that although there is a linear model that is more similar to the high-fidelity model than the low-fidelity model (i.e., convection-diffusion with anisotropic diffusivity and $k_r=0$ on the high-fidelity mesh) and thus would serve as a better initial guess, its existence is specific to our particular choice of models. The auxiliary and supplementary adjoint variables use a zero initial guess.

\Cref{tab:ref3D_newdata_QoI_diffmesh} shows the average QoI values, error estimates and solution times for the (non)linear solves. \Cref{tab:ref3D_newdata_times_diffmesh} shows the times needed to solve the inverse problems and to solve for the additional variables needed to obtain an error estimate. For six of the ten datasets, the high-fidelity inverse problem does not converge given the low-fidelity solution as an initial guess; these are solved using the $k_r=0$ solution as an initial guess so that true QoI errors can be calculated. The average high-fidelity inverse problem solution time shown in \Cref{tab:ref3D_newdata_times_diffmesh} includes only those cases (four of ten) that converged with the low-fidelity initial guess.
%
\begin{table}
\caption{Average QoI values and errors from solving inverse problem with mixed- and high-fidelity models and noisy data; relative errors are with respect to true high-fidelity QoI.}
\label{tab:ref3D_newdata_QoI_diffmesh}
\centering
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative  \\
& & & (Estimated) & (Actual) & Error (Actual) \\ \hline
LF   & 0    & 166774 & --    & -84326 & -102.3 \\
MF01 & 5.2  & 164597 & 4347  & -82149 & -99.65  \\
MF02 & 11.4 & 88867  & -5921 & -6418  & -7.79  \\
MF03 & 16.3 & 85237  & -2414 & -2789  & -3.38  \\
MF04 & 22.0 & 83411  & -724  & -963   & -1.17  \\
MF05 & 27.7 & 82664  & -18   & -216   & -0.26 \\
HF   & 100  & 82500  & --    & --     & --  \\ \hline
\end{tabular}
\end{table}

>>>>

\Cref{tab:ref3D_diffmesh} shows the error at the end of each adaptive iteration. Each iteration of the adaptive algorithm uses the solution of the previous iteration as its initial guess. The number of degrees of freedom of each of the primary (and, if applicable, auxiliary) variables at each iteration is also given; the supplementary adjoint is solved on the high-fidelity mesh and thus each of its variables has the same number of degrees of freedom as each of the primary variables in the high-fidelity inverse problem. The high-fidelity inverse problem does not converge when the low-fidelity solution is used as an initial guess. Instead, we solve the inverse problem for an intermediate model
%
\begin{equation}
\nabla\cdot(n\vec{V}u - nD\nabla u) + k_ru^2 = f(q)
\end{equation}
%
and use the solution as an initial guess for the high-fidelity problem; equivalently, we use two steps of natural continuation on the reaction parameter: $k_r=0$, then $k_r=4.2\cdot10^{-4}$.
%
\begin{table}[htbp]
\label{tab:ref3D_diffmesh}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{DOFs} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative \\
& & & & (Estimated) & (Actual) & Error (Actual)  \\ \hline
LF   & 0    & 9537  & 168710 & -16463 & -85663 & -103    \\
MF01 & 5.2  & 13417 & 167366 & -7207  & -84319 & -102    \\
MF02 & 11.4 & 17895 & 89777  & -6208  & -6730  & -8.10   \\
MF03 & 16.3 & 21001 & 85880  & -2473  & -2833  & -3.41   \\
MF04 & 22.0 & 24528 & 83902  & -711   & -855   & -1.03   \\
MF05 & 27.7 & 27984 & 83119  & 32     & -72    & -0.087  \\
HF   & 100  & 70785 & 83047  & --     & --     & --    \\ \hline
\end{tabular}
\caption{Runtime and relative errors of adaptive algorithm iterations given relative error tolerance of $0.1\%$; relative errors are given with respect to the high-fidelity QoI estimate.}
\end{table}
%

@@

To illustrate this offline-online application, we generate ten sets of noisy observations to represent the actual data gained during the online phase; the noisy observations are generated by taking the observations used in the adaptive algorithm and adding Gaussian white noise with standard deviation of $\sigma=0.02$ (equivalent to, on average, $5\%$ of the observed values). We then solve the inverse problem using each of the mixed-fidelity models depicted in \Cref{tab:ref3D_diffmesh} and \Cref{fig:divvy3D_diffmesh} as well as the high-fidelity model. The low-fidelity inverse problem is first solved and used as an initial guess for the higher-fidelity problems; we note that although there is a linear model that is more similar to the high-fidelity model than the low-fidelity model (i.e., convection-diffusion with anisotropic diffusivity and $k_r=0$ on the high-fidelity mesh) and thus would serve as a better initial guess, its existence is specific to our particular choice of models. The auxiliary and supplementary adjoint variables use a zero initial guess.

\Cref{tab:ref3D_newdata_QoI_diffmesh} shows the average QoI values, error estimates and solution times for the (non)linear solves. \Cref{tab:ref3D_newdata_times_diffmesh} shows the times needed to solve the inverse problems and to solve for the additional variables needed to obtain an error estimate. For six of the ten datasets, the high-fidelity inverse problem does not converge given the low-fidelity solution as an initial guess; these are solved using the $k_r=0$ solution as an initial guess so that true QoI errors can be calculated. The average high-fidelity inverse problem solution time shown in \Cref{tab:ref3D_newdata_times_diffmesh} includes only those cases (four of ten) that converged with the low-fidelity initial guess.
%
\begin{table}
\label{tab:ref3D_newdata_QoI_diffmesh}
\centering
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Case} & \multirow{2}{*}{$\%$HF} & \multirow{2}{*}{QoI} & Error & Error & $\%$ Relative  \\
& & & (Estimated) & (Actual) & Error (Actual) \\ \hline
LF   & 0    & 166774 & --    & -84326 & -102.3 \\
MF01 & 5.2  & 164597 & 4347  & -82149 & -99.65  \\
MF02 & 11.4 & 88867  & -5921 & -6418  & -7.79  \\
MF03 & 16.3 & 85237  & -2414 & -2789  & -3.38  \\
MF04 & 22.0 & 83411  & -724  & -963   & -1.17  \\
MF05 & 27.7 & 82664  & -18   & -216   & -0.26 \\
HF   & 100  & 82500  & --    & --     & --  \\ \hline
\end{tabular}
\caption{Average QoI values and errors from solving inverse problem with mixed- and high-fidelity models and noisy data; relative errors are with respect to true high-fidelity QoI.}
\end{table}
